Notes

### 2022-08-10 thesis/src/background-related-work.tex
Reorganize and start writing

% \documentclass[ba,preprint]{imsart}% use this for supplement article
\documentclass[ba]{imsart}
%
\pubyear{2022}
\volume{TBA}
\issue{TBA}
\doi{0000}
\arxiv{2010.00000}
\firstpage{1}
\lastpage{1}

%
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{bm}
\usepackage{natbib}
\usepackage[
  colorlinks,
  citecolor=blue,
  urlcolor=blue,
  filecolor=blue,
  backref=page
]{hyperref}
\usepackage{graphicx}

%% Comentarios
\usepackage{todonotes}
\usepackage{ulem}

\startlocaldefs
\numberwithin{equation}{section}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}{Proposition}[section]

\renewcommand{\epsilon}{\varepsilon}

\newcommand{\N}{\mathbb{N}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Hcal}{\ensuremath\mathcal{H}}
\newcommand{\Var}{\ensuremath\operatorname{Var}}
\definecolor{violet}{rgb}{0.7,0,0.6}

%% Scalar product
\newcommand\dotprod[2]{\left\langle #1, #2 \right\rangle}

%% Colors
\definecolor{comment-red}{RGB}{240,30,70}

%% Environment for colored comments
\newenvironment{comment}[1][comment-red]
{
\noindent \color{#1}
}
{
\color{black}
}

%% Inline comments
\newcommand\incomment[2][comment-red]{\color{#1}[\textit{#2}]\color{black}}
\endlocaldefs

\begin{document}

\begin{frontmatter}
\title{Bayesian RKHS-based methods in functional regression\support{Support information of the article.}}
\runtitle{Bayesian RKHS-based methods in functional regression}

\begin{aug}
\author{\fnms{First} \snm{Author}\thanksref{addr1,t1,t2,m1}\ead[label=e1]{first@somewhere.com}},
\author{\fnms{Second} \snm{Author}\thanksref{addr1,t3,m1,m2}\ead[label=e2]{second@somewhere.com}}
\and
\author{\fnms{Third} \snm{Author}\thanksref{addr2,t1,m2}%
\ead[label=e3]{third@somewhere.com}%
\ead[label=u1,url]{http://www.foo.com}}

\runauthor{F. Author et al.}

\address[addr1]{Address of the First and Second authors
     Usually a few lines long
    \printead{e1} % print email address of "e1"
    \printead*{e2}
}

\address[addr2]{Address of the Third author
    Usually a few lines long
    Usually a few lines long
    \printead{e3}
    \printead{u1}
}

\thankstext{t1}{Some comment}
\thankstext{t2}{First supporter of the project}
\thankstext{t3}{Second supporter of the project}

\end{aug}

\begin{abstract}
We propose a novel Bayesian approach for functional linear and logistic regression models based on the theory of Reproducing Kernel Hilbert Spaces (RKHS's). These new models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional approximation to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subspace of the RKHS of interest. By imposing a suitable prior distribution on this space, we can perform data-driven inference via standard Bayes methodology, and the posterior distribution can be estimated through Markov chain Monte Carlo methods. Several prediction strategies derived from this posterior distribution turn out to be competitive against other usual alternatives in both simulated examples and real data sets, including a Bayesian-motivated variable selection procedure.
\end{abstract}

\begin{keyword}[class=MSC]
\kwd[primary ]{62M20}
\kwd[; secondary ]{62F15}
\end{keyword}

\begin{keyword}
\kwd{functional data}
\kwd{linear regression}
\kwd{logistic regression}
\kwd{reproducing kernel Hilbert space}
\kwd{Bayesian inference}
\end{keyword}

\end{frontmatter}

\section{Introduction}\label{sec:intro}

The problem of predicting a scalar response from a functional covariate is one that has gained traction over the last few decades, as more and more data is being generated with an ever-increasing level of granularity in the measurements. While in principle the functional data could be simply regarded as a discretized vector in a very high dimension, there are often many advantages in taking into account the functional nature of the data, ranging from modeling the possibly high correlation among points that are close in the domain, to extracting information that may be hidden in the derivatives of the function in question. As a consequence, numerous proposals have arisen on how to suitably deal with functional data, all of them encompassed under the term Functional Data Analysis (FDA), which essentially explores statistical techniques to process, model and make inference on data varying over a continuum. A partial survey on such techniques and methods is \citet{cuevas2014partial}, while a more detailed exposition of the theory and applications can be found for example in \citet{hsing2015theoretical} or the book by \citet{horvath2012inference}.

FDA is undoubtedly an active area of research, which finds applications in a wide variety of fields, such as biomedicine, finance, meteorology or chemistry \citep[see for example][]{ullah2013applications}. Accordingly, there are many recent contributions on how to tackle functional data problems, both from a theoretical and practical standpoint. Chief among them is the approach of reducing the problem to a finite-dimensional one, for example using a truncated basis expansion or spline interpolation methods \citep[e.g.][]{muller2005generalized, aguilera2013comparative}. At the same time, much effort has also been put into the opposite task: generalizing different concepts to the infinite-dimensional framework. Examples of this endeavor include a functional Mahalanobis distance \citep{berrendero2020mahalanobis} or a functional partial least squares algorithm \citep{delaigle2012methodology}. Another technique found in the related literature is the use of Gaussian processes to model the functional behavior of the data \citep[see for instance][]{shi2011gaussian}. These ideas extend the theory of Gaussian process regression in classical finite-dimensional settings \citep{rasmussen2004gaussian}, providing an alternative Bayesian approach to functional inference and prediction problems. Additional non-parametric methods for functional prediction and classification were notably explored in \citet{ferraty2006nonparametric}.

In this work we are concerned with functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges as well as some subtle drawbacks that emerge as a result of working in infinite dimensions. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\). We will further suppose without loss of generality that \(X\) is centered, that is to say, its mean function \(m(t)=\E[X(t)]\) vanishes for all \(t\in[0,1]\). In addition, we will tacitly assume the existence of a \textit{labeled} data set \(\mathcal D_n =\{(X_i, Y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), and our aim will be to accurately predict the response corresponding to unlabeled samples from \(X\).

@@

A similar \(L^2\)-based functional logistic equation can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{eq:l2-logistic-model}
  \mathbb P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\beta}\}},
\end{equation}
where \(\alpha_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this situation, the most common way of estimating the slope function \(\beta\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but there is also the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions \citep[see][Sec.~3.2]{berrendero2018functional}.

@@

\subsubsection{Organization of the paper}

In Section~\ref{sec:rkhs} we review the basics of the theory of RKHS's from a probabilistic point of view. Section~\ref{sec:methodology} is devoted to explaining the Bayesian methodology and the functional regression models we propose. The empirical results of the experimentation are contained in Section~\ref{sec:results}, which also includes a short discussion of computational details. Lastly, the conclusions drawn from this work are presented in Section~\ref{sec:conclusion}.

\subsection{Some essentials on RKHS's and notation}\label{sec:rkhs}

The methodology proposed in this work relies heavily on the use of RKHS's, so before diving into it we will briefly describe the main characteristics of these spaces (for a more detailed account, see for example~\citet{berlinet2004reproducing}). Let us denote by \(K(t, s)= \mathbb E[X(t)X(s)]\) the covariance function of the centered process \(X\), and in what follows suppose that it is continuous. To construct the RKHS \(\Hcal(K)\) associated with the covariance function, we start by defining the functional space \(\Hcal_0(K)\) of all finite linear combinations of evaluations of \(K\), that is,
\begin{equation}\label{eq:h0}
\Hcal_0(K) = \left\{ f \in L^2[0,1]: \ f(\cdot) = \sum_{i=1}^p a_i K(t_i, \cdot), \ p \in \N, \ a_i \in \R, \ t_i \in [0, 1] \right\}.
\end{equation}
This space is endowed with the inner product \(\dotprod{f}{g}_K = \sum_{i, j} a_i b_j K(t_i, s_j)\), given that \(f(\cdot)=\sum_i a_i K(t_i, \cdot) \) and \(g(\cdot)=\sum_j b_j K(s_j, \cdot)\). Then, \(\Hcal(K)\) is defined to be the completion of \(\Hcal_0(K)\) under the norm induced by the scalar product \(\dotprod{\cdot}{\cdot}_K\). As it turns out, functions in this space satisfy the \textit{reproducing property} \(\dotprod{K(t, \cdot)}{f}_K = f(t)\) for all \(f \in \Hcal(K)\) and \(t \in [0, 1]\). An important consequence is that \(\Hcal(K)\) is a space of genuine functions and not of equivalence classes, since the values of the functions at particular points are in fact relevant, unlike in \(L^2\)-spaces.

@@

\section{Bayesian methodology for RKHS-based functional regression models}\label{sec:methodology}

In this section we present the precise models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration \citep[see][]{berrendero2018functional, berrendero2019rkhs} are those obtained by taking a functional parameter \(\alpha \in \Hcal(K)\) and replacing the scalar product for \(\dotprod{X}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}. For practical and computational reasons, the value of \(p\), the dimensionality of the model, will be fixed beforehand in a suitable way (see Section~\ref{sec:model-choice} for details), and we will regard \(\beta_j\) and \(t_j\) as free parameters. Thus, we will search for our functional parameter in the set
\begin{equation}\label{eq:h0p}
\Hcal_{0,p}(K)=\left\{ \sum_{j=1}^p \beta_j K(t_j, \cdot): \ \beta_j \in \R, \ t_j \in [0, 1]\right\}.
\end{equation}

The general idea will be to impose a prior distribution on these free parameters (and possibly others) to eventually derive a posterior model after incorporating the available sample information.  Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x(\alpha)\), where \(x=X(\omega)\) and \(\Psi_x\) is Loève's isometry. Hence, taking into account that \(\alpha \in \Hcal_{0,p}(K)\) and that \(\Psi_X(K(t, \cdot)) = X(t)\) by definition (see~\eqref{eq:loeves-isometry}), we can write \(\dotprod{x}{\alpha}_K \equiv \sum_j \beta_j x(t_j)\) when \(\alpha(\cdot)=\sum_j\beta_j K(t_j, \cdot)\).

In view of~\eqref{eq:h0p}, to set a prior distribution on the unknown function \(\alpha\) (that is, a prior distribution on the functional space \(\Hcal_{0,p}(K)\)) it suffices to consider \(p\)-dimensional continuous prior distributions for the coefficients \(\beta_j\) and the times \(t_j\) separately. Thanks to this parametric approach, the challenging task of setting a prior distribution on a space of functions is considerably simplified, while simultaneously not constraining the model to any specific distribution (in contrast to, for instance, Gaussian process regression methods). Moreover, note that starting from a probability distribution \(\mathbb{P}_0\) on \(\Hcal_0(K)\) we can obtain a probability distribution \(\mathbb{P}\) on \(\Hcal(K)\) merely by defining \(\mathbb{P}(B) = \mathbb{P}_0(B\cap \Hcal_0(K))\) for all Borel sets \(B\). Consequently, our simplifying assumption on \(\alpha\) is not very restrictive, since any prior distribution on \(\Hcal_0(K)\) can be directly extended to a prior distribution on \(\Hcal(K)\). Even though we actually work on \(\Hcal_{0,p}(K)\), the discrete parameter \(p\) can still be selected in several meaningful ways that make use of the available data, and the set of feasible values is not very large in practice.

In any case, after selecting a suitable prior distribution \(\pi(\theta)\) for the finite-dimensional parameter vector \(\theta\), we can resort to Bayes' theorem to perform the inference step, which in the case of i.i.d. samples amounts to
\begin{equation}\label{eq:bayes-theorem}
  \pi(\theta \mid \mathcal D_n) \propto \left( \prod_{i=1}^n \pi(Y_i\mid X_i, \theta) \right)\pi(\theta).
\end{equation}
In Sections~\ref{sec:rkhs-linear-model} and~\ref{sec:rkhs-logistic-model} we proceed to specify the parameter spaces, prior distributions and concrete models for \(\pi(Y | X,\theta)\) considered in the case of functional linear regression and functional logistic regression, respectively. Even though in~\eqref{eq:bayes-theorem} we have omitted the possibly intractable integral related to the normalizing constant, sampling from the (approximate) posterior distribution can be accomplished via Markov Chain Monte Carlo (MCMC) methods \citep[e.g.][]{brooks2011handbook}. Note that the use of MCMC algorithms introduces a source of stochasticity in the prediction procedure.

\subsection{Functional linear regression}\label{sec:rkhs-linear-model}

In the case of functional linear regression, the simplified RKHS model considered in this work is given by
\begin{equation}\label{eq:rkhs-model-linear}
  Y = \alpha_0 + \Psi_X(\alpha) + \epsilon = \alpha_0 + \sum_{j=1}^p \beta_j X(t_j) + \epsilon,
\end{equation}
where \(\alpha(\cdot)=\sum_{j=1}^p\beta_jK(t_j, \cdot) \in \Hcal_{0,p}(K), \beta_j \in \R\), \(t_j \in [0, 1]\), \(\alpha_0\in\R\), and \(\epsilon\) is a normally-distributed error term independent from \(X\), with \(\E[\epsilon]=0\) and \(\Var(\epsilon)=\sigma^2\in \R^+\). This model is essentially a finite-dimensional approximation from a functional perspective to the more general RKHS model that assumes \(\alpha \in \Hcal(K)\), proposed in~\citet{berrendero2019rkhs}.

When \(p\) is fixed, the parameter space of dimension \(2p + 2\) becomes \(\Theta_p = \R^p \times [0, 1]^p \times \R \times \R^+\), and in the sequel a generic element of this space will be denoted by \(\theta = (\beta_1,\dots, \beta_p, t_1,\dots, t_p, \alpha_0, \sigma^2) \equiv (b, \tau, \alpha_0, \sigma^2)\). Before proceeding any further, observe that we can rewrite model~\eqref{eq:rkhs-model-linear} in a more explicit and practical fashion in terms of the available sample information in \(\mathcal D_n\). For \(\theta \in \Theta_p\), the reinterpreted model assumes the form
\begin{equation}\label{eq:rkhs-model-linear-2}
  Y_i \mid X_i, \theta \ \stackrel{\text{i.i.d.}}{\sim} \mathcal N\left(\alpha_0 + \sum_{j=1}^p \beta_j X_i(t_j), \ \sigma^2\right), \quad i =1,\dots, n.
\end{equation}

It is worth mentioning that the model remains linear in the sense that it fundamentally involves a random variable belonging to the linear span of the process \(X\) in \(L^2(\Omega)\). Also, note that given the time instants \(t_j\), the model becomes a multiple linear model with the \(X(t_j)\) as scalar covariates. As a matter of fact, this RKHS model is particularly suited as a basis for variable selection methods, and also entails the classical \(L^2\)-model~\eqref{eq:l2-linear-model} under certain conditions \citep[see][Sec.~3]{berrendero2020general}. In addition, this model could be easily extended to the case of several covariates via an expression of type \(Y=\alpha_0 + \Psi_{X^{1}}(\alpha_1) + \cdots + \Psi_{X^{q}}(\alpha_q) + \epsilon\). In that case, as argued in \citet{grollemund2019bayesian} for a similar situation, if we were to set a prior distribution on all the parameters involved, we could recover the full posterior by looking alternately at the posterior distribution of each covariate conditional on the rest of them.

\subsubsection{The bayesian approach: prior selection and posterior derivation}

The prior distribution suggested for the parameter vector \(\theta \in \Theta_p\) is given by
\begin{align}\label{eq:prior-linear}
  \begin{split}
  \pi(\alpha_0, \sigma^2)              & \propto 1/\sigma^2,                                                     \\
  \tau                     & \sim \mathcal U([0, 1]^p),                                              \\
  b\mid \tau, \sigma^2 & \sim \mathcal N_p(b_0, g\sigma^2{\underbrace{\left(\mathcal X_\tau' \mathcal X_\tau + \eta I\right)}_{G_\tau}}^{-1}),
\end{split}
\end{align}
where \(I\) is the identity matrix, \(\mathcal X_\tau\) is the data matrix \((X_i(t_j))_{i,j}\), and \(b_0\in \R^p, \ g \in \R\) and \(\eta \in \R^+\) are hyperparameters of the model. On the one hand, note the use of a joint prior distribution on \(\alpha_0\) and \(\sigma^2\), which is a widely used non-informative prior known as Jeffrey's prior \citep{jeffreys1946invariant}. In any event, the estimation of \(\alpha_0=\E[Y]\) is straightforward, so it could have been left out of the model altogether. On the other hand, the prior on \(b\) is a slight modification of the well-known Zellner's g-prior \citep{zellner1986assessing}, in which a regularizing term is added to avoid ill-conditioning problems in the Gram matrix, obtaining a ridge-like Zellner prior controlled by the tuning parameter \(\eta\) \citep{baragatti2012study}. All in all, with a slight abuse of notation the proposed prior distribution becomes \(\pi(\theta) = \pi(b| \tau, \sigma^2)\pi(\tau)\pi(\alpha_0, \sigma^2)\).

As for the posterior distribution, we only compute a function proportional to its log-density, since that is all that is needed for the MCMC algorithm to work. A standard algebraic manipulation in~\eqref{eq:bayes-theorem} yields the following result.

\begin{proposition}
Under the linear model~\eqref{eq:rkhs-model-linear-2}, the prior distribution implied in~\eqref{eq:prior-linear} produces the log-posterior distribution
\begin{align*}
\log \pi(\theta \mid \mathcal D_n) \propto {} & \frac{1}{2\sigma^2}\left(\|\bm Y- \alpha_0\bm{1} - \mathcal X_\tau b\|^2 + \frac{1}{g}(b - b_0)'G_\tau(b - b_0) \right)\\
& + (p+n+2)\log\sigma - \frac{1}{2}\log |G_\tau|,
\end{align*}
where \(\bm Y=(Y_1,\dots,Y_n)'\) and \(\bm{1}\) is an \(n\)-dimensional vector of ones.
\end{proposition}

\subsubsection{Making predictions}

In order to generate predictions, let us recall that when performing the empirical posterior approximation, on each of the \(M\) steps of the iterative MCMC algorithm we get an approximate sample \(\theta^{(m)*}=(b^{(m)*}, \tau^{(m)*}, \alpha_0^{(m)*}, (\sigma^2)^{(m)*})\) of the posterior distribution \(\pi(\theta| \mathcal D_n)\). Assuming now a previously unseen test set \(\mathcal D'_{n'}\) in the same conditions as \(\mathcal D_n\), we propose to construct three different kinds of predictors based on the MCMC samples, each of them following a different strategy.

\begin{description}
  \item[Summarize-then-predict.] If we consider a point-estimate statistic \(T\) to act as a summary of the posterior distribution, we can get the corresponding estimates \(\hat{\theta}=(\hat b, \hat \tau, \hat{\alpha}_0, \hat{\sigma}^2) = T\{\theta^{(m)*}\} \equiv (T\{b^{(m)*}\}, T\{\tau^{(m)*}\}, T\{\alpha_0^{(m)*}\}, T\{(\sigma^2)^{(m)*}\})\) and predict the responses in the usual way following model \eqref{eq:rkhs-model-linear}, i.e.:
  \[
    \hat Y_i = \hat{\alpha}_0 + \sum_{j=1}^p \hat{\beta}_j X_i(\hat{t}_j), \quad i=1,\dots, n'.
  \]
  Note that in this case the variance \(\sigma^2\) is treated as a nuisance parameter. Although it contributes to measure the uncertainty in the approximations, its estimates are discarded in the final prediction.

  \item[Predict-then-summarize.] Alternatively, we can  look at the approximate posterior distribution as a whole and compute the predictive distribution of the simulated responses at each step of the chain following model~\eqref{eq:rkhs-model-linear-2}:
  \begin{equation}\label{eq:sampled-response-vector}
  \bm Y^{(m)*} := \left\{Y_i^{(m)*} \equiv Y_i \mid X_i, \theta^{(m)*}:\ i=1,\dots,n'\right\}, \quad m=1,\dots,M.
  \end{equation}
  Then, we can take the mean of all such simulated responses as a proxy for each response variable, that is,
  \[
  \hat Y_i = \frac{1}{M}\sum_{m=1}^M Y_i^{(m)*}, \quad i=1,\dots,n'.
  \]
  This method differs from the previous one in that it takes into account the full approximate posterior distribution instead of summarizing it directly.

  \item[Variable selection.] Lastly, we can focus only on the marginal posterior distribution of \(\tau|\mathcal D_n\) and select \(p\) time instants using a point-estimate statistic \(T\) as in our first strategy, but discarding the rest of the parameters. Specifically, we can consider the times \(\hat t_j = T\{t_j^{(m)*}\}\) and reduce the original data set to just the \(n\times p\) real matrix \(\{X_i(\hat t_j): i=1, \dots,n, \ j=1,\dots,p\}\). After this variable selection has been carried out, we can tackle the problem using a finite-dimensional linear regression model and apply any of the well-known prediction algorithms suited for this situation.
\end{description}

Note that these predictors can be obtained all at once after only one round of training (that is, an individual MCMC run to approximate the posterior distribution). As a consequence, what we have in practice is a single algorithm that can produce multiple predictors at the same computational cost, so that any of them can be chosen (or even switched back and forth) depending on the particularities of the problem at hand. Moreover, one could even contemplate an \textit{ensemble model} in which some kind of aggregation of several of the available prediction methods is performed to produce a final result.

\subsection{Functional logistic regression}\label{sec:rkhs-logistic-model}

In the case of functional logistic regression, we regard the binary response variable \(Y\in\{0, 1\}\) as a Bernoulli random variable given the regressor \(X=x \in L^2[0, 1]\), and suppose that \(\log\left(p(x)/(1-p(x))\right)\) is linear in \(x\), where \(p(x)=\mathbb P(Y=1| X=x)\). Then, following the approach  suggested by \citet{berrendero2018functional}, an RKHS model might be given  by the  equation
\begin{equation}\label{eq:rkhs-model-logistic}
  \mathbb P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \Psi_X(\alpha)\}}, \quad \alpha_0 \in \R, \ \alpha \in \Hcal_{0,p}(K).
\end{equation}

Indeed, note that this can be seen as a finite-dimensional approximation (but, still, with a functional interpretation) to the general RKHS functional logistic model proposed by these authors, which can be obtained by replacing \(\Hcal_{0,p}(K)\) with the whole RKHS space \(\Hcal(K)\). Now, if we aim at a classification problem, our strategy will be similar to that followed in the functional linear model: after incorporating the sample information, we can rewrite~\eqref{eq:rkhs-model-logistic} as
\begin{equation}\label{eq:rkhs-model-logistic-2}
Y_i \mid X_i,\theta \ \stackrel{\text{i.i.d.}}{\sim} \operatorname{Bernoulli}(p_i), \quad i=1,\dots, n,
\end{equation}
with
\begin{equation}\label{eq:rkhs-model-logistic-2-parameter}
  p_i = \mathbb P(Y_i=1 \mid X_i,\theta) = \frac{1}{\displaystyle 1 + \exp\left\{-\alpha_0 - \sum_{j=1}^p \beta_j X_i(t_j)\right\}}, \quad i=1,\dots, n,
\end{equation}
where in turn \(\beta_j\in\R\) and \(t_j\in[0, 1]\).

In much the same way as the linear regression model described above, this RKHS-based logistic regression model offers some advantages over the \(L^2\)-model. First and foremost, it has a more straightforward interpretation and allows for a workable Bayesian approach, as we will demonstrate below. Secondly, it can be shown that under mild conditions the general RKHS logistic functional model holds whenever the conditional distributions \(X | Y=i\) (\(i=0,1\)) are homoscedastic Gaussian processes~\citep[see Theorem 1 in][]{berrendero2018functional}; this provides a sound theoretical motivation for such model. Furthermore, a maximum likelihood approach for parameter estimation (although not considered here) is possible as well. Indeed, the use of a finite-dimensional approximation  mitigates the problem of non-existence of the MLE in the functional case. However, let us recall that even in finite-dimensional settings there are cases of quasi-complete separation in which the MLE does not exist \citep{albert1984existence}, though this issue can be circumvented using, for example, Firth's corrected estimator \citep{firth1993bias}.
\color{black}

\subsubsection{The Bayesian approach: prior selection and posterior derivation}

As far as prior distributions go, we propose to use the same ones as we did in~\eqref{eq:prior-linear} for the linear regression model. However, in this case the parameter \(\sigma^2\) is a nuisance parameter and can be safely ignored at the end, since it does not appear in the final model. The posterior distribution is then derived after a routine calculation.

\begin{proposition}
Under the logistic model~\eqref{eq:rkhs-model-logistic-2}, the prior distribution implied in~\eqref{eq:prior-linear} produces the log-posterior distribution
\begin{align*}
  \log \pi(\theta \mid \mathcal D_n) \propto {} & \sum_{i=1}^n \left[ \left(\alpha_0 + \Psi_{X_i}(\alpha)\right)Y_i - \log\left(1 + \exp\left\{\alpha_0 + \Psi_{X_i}(\alpha)\right\}\right)\right]\\
  \quad &+ \frac{1}{2}\log |G_\tau| - (p+2)\log \sigma -\frac{1}{2g\sigma^2} (b - b_0)'G_\tau(b - b_0).
\end{align*}
Remember that \(\Psi_{X_i}(\alpha) = \sum_{j=1}^p \beta_j X_i(t_j)\).
\end{proposition}

\subsubsection{Making predictions}

Bear in mind that in this case we are essentially approximating the probabilities \(p_i\) in~\eqref{eq:rkhs-model-logistic-2-parameter}, so before producing a response we need to transform the predicted values to a binary output in \(\{0, 1\}\). According to the usual criterion of minimizing the misclassification probability, the Bayes optimal rule is recovered by predicting \(\hat Y=1\) whenever \(\mathbb P(Y=1|X) \geq 1/2\). Nevertheless, for a more general cost function one could consider other criteria that would lead to evaluating whether \(\mathbb P(Y=1|X) \geq \gamma\) for some threshold \(\gamma\in[0, 1]\).

With this last strategy in mind, the summarize-then-predict approach on the approximate posterior distribution is analogous to the linear regression case:
\[
\hat Y_i = \mathbb I \left( \left[\displaystyle 1 + \exp\left\{-\hat\alpha_0 - \sum_{j=1}^p \hat\beta_j X_i(\hat t_j)\right\}\right]^{-1} \geq \gamma \right), \quad i=1,\dots,n',
\]
where \(\mathbb I\) is the indicator function (\(\mathbb I(P)\) is \(1\) if \(P\) is true and \(0\) otherwise). The hat estimates are obtained once again through a summary statistic \(T\) of the corresponding marginal posterior distributions.

On the other hand, the prediction method that takes into account the entire posterior approximation (i.e. the predict-then-summarize approach) is somewhat different now, since there is the question of which response (the Bernoulli variables in~\eqref{eq:rkhs-model-logistic-2} or the raw probabilities in~\eqref{eq:rkhs-model-logistic-2-parameter}) to consider when averaging the posterior samples. Hence, there are primarily two possible outcomes:
\begin{description}
  \item[Average sampled probability.] If we choose to average the approximate probabilities \(p_i^{(m)*} = \mathbb P(Y_i =1 | X_i,\theta^{(m)*})\) computed following~\eqref{eq:rkhs-model-logistic-2-parameter}, the resulting predictor is
  \[
    \hat Y_i = \mathbb I\left(\frac{1}{M} \sum_{m=1}^M p_i^{(m)*} \geq \gamma\right), \quad i=1,\dots,n'.
  \]
  \item[Average sampled response.] Deciding to average the approximate binary responses \(Y_i^{(m)*}\) instead (see~\eqref{eq:sampled-response-vector}) leads to computing the predictions as
\[
  \hat Y_i = \mathbb I\left(\frac{1}{M} \sum_{m=1}^M Y_i^{(m)*} \geq \gamma\right), \quad i=1,\dots,n'.
\]
In this case, each \(Y_i^{(m)*}\) follows a Bernoulli distribution of parameter \(p_i^{(m)*}\), for \(m=1,\dots,M\). Note that when \(\gamma=1/2\) this is equivalent to predicting \(Y_i\) from the majority vote of all the \(Y_i^{(m)*}\).
\end{description}

Lastly, the variable selection method is essentially the same as in the case of linear regression: we select \(p\) time instants from each trajectory based on a summary of the posterior distribution \(\tau | \mathcal D_n\), and then feed the reduced data set to a finite-dimensional binary classification procedure.

\subsection{Model choice and validation}\label{sec:model-choice}

In this section we gather together several remarks on the main choices made throughout the design of our model, as well as some validation strategies that attempt to measure the goodness-of-fit of the model given the observed data.

\paragraph{The choice of \(\bm p\).} One of the key decisions in our Bayesian modeling scheme was whether to consider the number of components \(p\) as a member of the parameter space and integrate it into the model. While theoretically we could impose a prior distribution on \(p\) as well (e.g. a categorical distribution with a fixed maximum value), we found that it would have some unwanted practical implications. For instance, it would make the implementation more complex, since the dimensionality of the parameters \(b\) and \(\tau\) would need to be fixed at a certain maximum value beforehand, but the working value of \(p\) within the MCMC algorithm would vary from one iteration to the next. In this case we would face a label switching issue \citep[c.f.][Sec.~2.3]{grollemund2019bayesian}, as we would have no way of knowing which set of parameters is ``active'' at any given time. Another possibility would be to adapt a purely Bayesian model selection technique \citep[e.g.][Ch.~7]{piironen2017comparison, gelman2013bayesian} to our framework, or even derive some model aggregation methods to combine the posterior distributions obtained for different-sized models. In the end, for the sake of simplicity we decided to let \(p\) be an hyperparameter, so that we could use any model selection criteria (e.g. BIC, DIC, cross-validation, \ldots) to select its optimal value. This is equivalent to imposing a degenerate prior distribution on \(p\). Moreover, the experiments carried out indicate that even low values of \(p\) provide sufficient flexibility in most scenarios (see Section~\ref{sec:results}). \incomment{Reversible MCMC}

\begin{comment}
LO QUE NO SIRVA, GUARDAR PARA EL TFM.

Reescribir label switching (nos pasa; lo que pasa que con p free se acentua).

Ocurre porque el likelihood es simetrico: L(theta mid y) = L(thetaperm mid y) para cualquier permutacion (hay p! posibles) --> no-identificabilidad.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Copyright (c) 2022 Antonio Coín Castro
%
% This work is licensed under a
% Creative Commons Attribution-ShareAlike 4.0 International License.
%
% You should have received a copy of the license along with this
% work. If not, see <http://creativecommons.org/licenses/by-sa/4.0/>.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Background and related work}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A similar \(L^2\)-based functional logistic equation can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{eq:l2-logistic-model}
  \Prob(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\beta}\}},
\end{equation}
where \(\alpha_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this situation, the most common way of estimating the slope function \(\beta\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but there is also the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions \citep[see][Sec.~3.2]{berrendero2018functional}.

@@

\section{Reproducing kernel Hilbert spaces}\label{sec:rkhs}

For a more detailed account, see for example~\citet{berlinet2004reproducing}.

Suppose \(X=X(t)\) is a \(L^2\)-stochastic process with trajectories in \(L^2[0, 1]\), and without loss of generality we can assume \(\E[X(t)]=0\) for all \(t\in[0,1]\). Let us denote by \(K(t, s)= \E[X(t)X(s)]\) the covariance function of the process \(X\), and in what follows suppose that it is continuous. To construct the RKHS \(\Hcal(K)\) associated with the covariance function, we start by defining the functional space \(\Hcal_0(K)\) of all finite linear combinations of evaluations of \(K\), that is,
\begin{equation}\label{eq:h0}
\Hcal_0(K) = \left\{ f: \ f(\cdot) = \sum_{i=1}^p a_i K(t_i, \cdot), \ p \in \N, \ a_i \in \R, \ t_i \in [0, 1] \right\}.
\end{equation}
Note that, as subsets, \(\Hcal_0(K)\subset L^2[0,1]\). However, this space can be endowed with an inner product different from the one induced by \(L^2[0,1]\), namely \(\dotprod{f}{g}_K = \sum_{i, j} a_i b_j K(t_i, s_j)\) for \(f(\cdot)=\sum_i a_i K(t_i, \cdot) \) and \(g(\cdot)=\sum_j b_j K(s_j, \cdot)\).

\begin{proposition} \(\dotprod{\cdot}{\cdot}_K\) is an inner product.
  \begin{proof}
    Symmetry and linearity are clear from the definition, while non-negativity is a direct consequence of \(K\) being a covariance function, and thus positive semidefinite.
  \end{proof}
\end{proposition}

Then, \(\Hcal(K)\) is defined to be the completion of \(\Hcal_0(K)\) under the norm induced by the scalar product \(\dotprod{\cdot}{\cdot}_K\), turning it into a genuine Hilbert space. As it turns out, functions in this space satisfy the so-called \textit{reproducing property}, which can be easily checked from the definitions.

\begin{proposition}
  If \(f\in \Hcal(K)\), then \(\dotprod{K(t, \cdot)}{f}_K = f(t)\) for all \(t \in [0, 1]\).
  \begin{proof}

\end{proof}
\end{proposition}

This is where the name ``reproducing kernel'' comes from. An important consequence is that \(\Hcal(K)\) is a space of genuine functions and not of equivalence classes, since the values of the functions at particular points are in fact relevant, unlike in \(L^2\)-spaces.

\begin{outcomment}
  A kernel defines an RKHS and vice-versa.
\end{outcomment}

@@
